---
title: "Part 1: Exploring Gapminder Data"
author: "Ben Brown"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(geonames)
library(gapminder)
library(countrycode)
```

### 1. Get the country information using geonames - remember there is a R package for this (see the lecture notes)

``` {r get-country-info, message = FALSE}
# Note that I edited my .Rprofile file to set geonamesUsername

countryInfo <- tibble(GNcountryInfo())
```
### 2. Merge gapminder and the country information from geonames
* Use left_join from dplyr to merge the tables
* Note that you cannot directly do this - gapminder writes the name of countries differently from geonames. To complete the merge, you need a unique key to match observations between the data frames
* There is neat little package for R called countrycode that helps solve this problem. countrycode() takes as an input a country’s name in a specific format and outputs it using whatever format you specify.
  - `gapminder` stores them using the country.name format
  - `geonames` stores them under the countryCode column using the iso2c format
  - I leave it to you to make the joining operation work
  
``` {r merge}
# change gapminder's country names to corresponding iso2c codes,
# Then merge with countryInfo
countryData <- gapminder %>%
  mutate(country = countrycode(sourcevar = country,
                               origin = "country.name",
                               destination = "iso2c")) %>%
  left_join(countryInfo, by = c("country" = "countryCode")) %>%
  mutate(areaInSqKm = as.numeric(areaInSqKm)) # Convert area to numeric

countryData
```

### 3. Calculate the population density for each observation

``` {r pop-density}

pop_density <- countryData %>%
  mutate(popDensity = pop / areaInSqKm) # Use gapminder's population statistic

pop_density

```
### 4. Produce an updated graph using population density
* If you want to be real fancy, estimate a statistical model or compare the relationship across continents

``` {r graph, message = FALSE}

ggplot(
  data = pop_density,
  mapping = aes(x = popDensity, y = lifeExp)
) + 
  geom_point(alpha = 0.2) +
  geom_smooth() +
  scale_x_log10() +
  labs(
    title = "As population density increases, life expectancy increases as well",
    x = "Population density (people per square km)",
    y = "Average life expectancy"
  )

```

The new graph shows that as population density increases, life expectancy does as well. What's interesting about both graphs is that there are clear lines of dots that have an upward trend. These lines correspond to one country over time, and they show that over time, countries' population, population density, and average life span have generally increased. It would thus be interesting to compare the relation between different countries at the same time or the same country across different time, to see if the year or the country matters more.

``` {r by-continent, message = FALSE}

ggplot(
  data = pop_density,
  mapping = aes(x = popDensity, y = lifeExp, color = continentName)
) + 
  geom_point(alpha = 0.2) +
  geom_smooth() +
  scale_x_log10() +
  # Use gapminder's colors for the different continents
  scale_color_discrete(name = NULL) +
  labs(
    title = "Population density versus life expectancy by continent",
    x = "Population density (people per square km)",
    y = "Average life expectancy"
  )

```

The continent where the correlation is most positive is South America. Asia and Africa also have a generally positive correlation, with a somewhat negative one at lower population densities. Europe and North America have less correlation than the other groups. Finally, Oceania's graph looks like there is some kind of sinusoidal correlation at first, but it is actually made up of two groups of points, each with a positive correlation, so there is likely positive correlation here as well.

It seems that less developed continents have a more positive relationship between population density and life expectancy. This is likely because over the past few decades, these areas had the highest population growth and the most health improvements, leading to older data points generally having both lower density and life expectancy, while newer data points had higher density and life expectancy. 
